---
  title: "Computational Musicology"
  author: "Melle van der Brugge"
  output:
    flexdashboard::flex_dashboard:
      orientation: columns
      vertical_layout: fill
      storyboard: true
      theme: 
        bg: "#f4b41a"
        fg: "#143d59" 
      social: menu
      source_code: embed
      navbar: 
      - { title: "Link to playlist", href: "https://open.spotify.com/playlist/37i9dQZF1DX76Wlfdnj7AP", align: right, icon: glyphicon}
      self_contained: false
---

```{r setup}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(spotifyr)
library(ggplot2)
library(shiny)
library(compmus)
library(knitr)
library(tidymodels)
library(heatmaply)
library(ggdendro)
library(ggpubr)
```

# The Playlist {data-icon="fa-music"}

**Get your Beast Mode on**!\
"*Beast mode refers to the aggressive, animalistic persona that one might assume when in competition or combat in order to overpower an opponent*."

The corpus I chose for Computational Musicology is playlists labelled "Workout". During covid, I often went to the gym to get out of the house. Furthermore, I always took my phone and Air Pods if I went to the gym. Music was an essential part of my gym workout. Moreover, I always listen to the label "Workout" on Spotify during my sessions. So, what does it mean to be a "workout" playlist? It interests me that it needs to give you energy because you are exercising, and you do not want to stop. It needs to help you focus on your movements. And you don't want to get bored or feel the pain from the exercise. And does all the workout music have lyrics to keep yourself motivated, or is this not necessary. One should consider the emotions, memories, and associations that different songs evoke. For some people, the extent to which they identify with the singer's emotional state and viewpoint determines how motivated they feel

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/playlist/37i9dQZF1DX76Wlfdnj7AP?utm_source=generator" width="100%" height="380" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture">

</iframe>

It can be hard to stay motivated during a workout, especially when you have so many other things on your mind. However, exercise can help distract you from your to-do lists and everyday anxieties, something we all could use during the novel coronavirus age. \
"We know about the mental and physical benefits of exercise for the brain, and that it aids with sleep," says Jordan Metzl, MD, a sports medicine physician at Hospital for Special Surgery in New York City. "When people are ruminating about something, exercise helps." \
If you are struggling to stay in the zone, you can do a couple of things to amplify your sweat sesh. Lacing up a fresh pair of sneakers, for starters. Or a crowd favorite: cranking the beats.

The right workout playlist can make a huge difference in how you feel and perform during your workouts. Listening to high-tempo music (the kind that's around 170 heartbeats per minute) made people perceive their workouts as more effortless, even when they were challenging, and it boosted cardiovascular benefits more than lower tempo tunes, according to a recent study published in Frontiers in Psychology. \
What is more: Music may change how you respond to fatigue---and even boost your endurance, per research published in the Journal of Sports Exercise Psychology. Plus, other research from Sage Journal has shown that tunes can make you feel more powerful and that people who listen to music they deemed "pleasing" had higher levels of serotonin, also known as the "feel-good" hormone, according to findings published in the book Progress in Brain Research. Combine that with a workout, which already amps up your endorphins, and you will really be treating your body right (Jones & Jones, 2021).

But how does a workout playlist distinguish itself from other playlists? And what are the characteristics of a workout playlist? Check out the storyboard and find out!


# Storyboard {.storyboard data-icon="fa-tachometer"}


### **Exploring the playlist:** Understanding the Beast Mode playlist using Spotify audio features

```{r exploring}
beast_mode <-
  get_playlist_audio_features("", "37i9dQZF1DX76Wlfdnj7AP")

feature_names <- names(beast_mode)[6:16]

beast_mode %>%
  select(c('playlist_name', feature_names)) %>%
  pivot_longer(cols = feature_names) %>%
  ggplot(aes(x = value)) +
  geom_density(aes(color = playlist_name), alpha = 0.5) +
  facet_wrap(~name, ncol = 3, scales = 'free') +
  labs(title = 'Spotify Audio Feature Density',
       x = '', y = 'density') +
  theme(axis.text.y = element_blank()) 


```

------------------------------------------------------------------------

When your mission is to create the perfect workout playlist, the first thing you want is to have songs you enjoy, of course. However, there are some other features you want your songs to have in your workout playlist. You want energising tracks that will keep you moving. You maybe want songs that have inspiring and motivational lyrics. And you should consider the beat; working out in time to the beat can help you get in your " zone " to make the most of your workout.

A few features jump out in the figure on the left.\
The danceability of the playlist is around 0.74. A value of 0.0 is the least danceable, and 1.0 is the most danceable. Danceability describes how suitable a track is for dancing based on a combination of musical elements, including tempo, rhythm stability, beat strength, and overall regularity. You also want this in your workout playlist to keep the tempo, energy in your workout

Energy is also high in the Beast mode playlist. Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. When doing your workout, you want to have the energy to finish all your exercises. Songs having high energy will help you achieve this.

The instrumentalness is very low in the Beast mode playlist. So Spotify predicts that the tracks do contain vocal content. And this is something you want in your playlist, lyrics for inspiring and motivation, but also lovely beats to keep your energy

The speechiness is mostly under 0.33, representing music and other non-speech-like tracks. This is an exciting insight into the corpus because inspiring and motivational lyrics could help you work out. People who listen to the Beast mode playlist prefer danceability and energy over inspiring and motivational lyrics.

The loudness of the playlist has an average of around -5. The overall loudness of a track is in decibels (dB). Loudness values are averaged across the entire track and help compare the relative loudness of tracks. Loudness is the sound quality that is the primary psychological correlate of physical strength (amplitude). Values typically range between -60 and 0 db.\
Because the range is between -60 and 0,  -5 is relatively high. But why does loudness help in your workout? The volume itself is not necessarily a contributing factor toward the level of enjoyment. It is more that the louder that music is, the quieter everything else seems in comparison. One can literally drown out daily life, the daily grind, personal emotional issues etc, all seem to periodically 'go away' whilst listening to loud music. It becomes a focus point; you can entirely focus on your workout. There are no background noises that interrupt the flow of your gym session.

If you want to know the meaning of all Spotify audio features, check out the track's audio features: <https://developer.spotify.com/documentation/web-api/reference/#/operations/get-audio-features>

### **The Workout Feeling:** Which emotions are conveyed by the songs in Beast Mode

```{r feeling}
beast_mode <-
  get_playlist_audio_features("", "37i9dQZF1DX76Wlfdnj7AP")

emotionalQuadrant <- ggplot(data = beast_mode, aes(x = valence, y = energy, color = "red")) +
  geom_jitter() +
  geom_vline(xintercept = 0.5) +
  geom_hline(yintercept = 0.5) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 1)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  annotate('text', 0.25 / 2, 0.95, label = "Angry / Turbulent") +
  annotate('text', 1.75 / 2, 0.95, label = "Joyful / Happy") +
  annotate('text', 1.75 / 2, 0.05, label = "Peace / Chill") +
  annotate('text', 0.25 / 2, 0.05, label = "Depressing / Sad") +
  labs(x= "Valence", y= "Energy") +
  ggtitle("Emotional quadrant Beast mode", "Based on energy y valence")  
emotionalQuadrant
```

------------------------------------------------------------------------

In the figure "Emotional quadrant Beast mode", you can see what feeling the Beast mode playlist gives. Do songs convey peace, joy, anger, and sadness? We can answer this question by looking at two measures Spotify has established: Valence and Energy.

Defined by Spotify, valence defines it as follows: A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry). On the other hand, energy defines as follows: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.

You can create a quadrant with a scatter plot that reveals this information with these two measurements. The Beast mode playlist is high in energy, and valence has a nice distribution.

### **The Remix Effect:** What is the difference in tempo between a "normal" version and and remixed version

```{r trust nobody}
trust_nobody_normal <- get_tidy_audio_analysis("08bDGEzsYd9ebCc7p23X4j")
      
normal <- trust_nobody_normal %>%
  tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) %>%
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

trust_nobody_remix_valentino_khan <- get_tidy_audio_analysis("52wqMbmCqeTw37jd4zFAWr")

remix <- trust_nobody_remix_valentino_khan %>%
  tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) %>%
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

ggarrange(normal, remix, 
          labels = c("Normal Version", "Remix"),
          ncol = 1, nrow = 2)

```

------------------------------------------------------------------------

Here you can see two tempograms, both of the songs "Trust Nobody". Except one is the "normal" version of the track, and the other is the remixed version of the track. The remix version is in the Beast mode playlist; the normal version is not.

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/album/3wRs5NGdWArDIHoUw9UNMJ?utm_source=generator" width="100%" height="80" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture">

</iframe>

You can see that the song's beat is inconsistent until around 30 seconds in the song. After 30 seconds, a line is recognizable at around 150 BPM. The song starts slower, building up to the beat drop. This repeats after around 105 seconds. So it is clearly visible that the song has two slower parts, both around 30 seconds. This can be why the remixed version is in the Beast mode playlist, not the normal version. 

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/album/6r58Yml4mD6I4MXwN4UNkx?utm_source=generator" width="100%" height="80" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture">

</iframe>

In contrast to the normal version, the Valentino Khan version of Trust Nobody has a clear line almost the whole song at around 157 BPM. The BPM is even higher than the normal version. However, the biggest difference is missing the two slower parts in the remix version. The beat is the whole song very high. Furthermore, this can be why this song is in the Beast mode playlist. Earlier studies reported that listening to music with a fast tempo caused an increase in the heart rate (HR) and respiratory rate (Gomez & Danuser, 2007). And getting your heart to beat faster helps your body move oxygen and blood to your muscles (Blackburn, 2019), helping you in your workout!



### **Structure:** Why is 'Do It To It' the most popular song

```{r self similarity matrices}
	
do_it_to_it <-
  get_tidy_audio_analysis("20on25jryn53hWghthWWW3") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  )
bind_rows(
  do_it_to_it %>% 
    compmus_self_similarity(pitches, "aitchison") %>% 
    mutate(d = d / max(d), type = "Chroma"),
  do_it_to_it %>% 
    compmus_self_similarity(timbre, "euclidean") %>% 
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>% 
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(option = "E", guide = "none") +
  theme_classic() + 
  labs(x = "", y = "Do It To It")
```

***

On the left, the two self-similarity matrices show pitch- and timbre-based self-similarity within the song "Do it to it" by ACRAZE and Cherish. Do It To It is the most popular track in the Beast Mode playlist. However, can we understand the song's structure and conclude why it is the most popular song and a good workout song? 

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/track/20on25jryn53hWghthWWW3?utm_source=generator" width="100%" height="80" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture"></iframe>

Both the chrome matrix and the timbre matrix have a checkerboard-like pattern. This is caused by an alternation between the introduction, the chorus, a bridge, the chorus again and the outro. There is no significant variation in the timbre matrix; this could be because it is instrumental, almost the same for the whole song. The timbre and chroma matrices also show that the song is pretty balanced. \ 
You can even see a little "build-up" in the first chorus just before the bridge, starting at around 65 seconds until 75 seconds when the bridge starts. 


### **Take it to a higher tempo** What are the components of the 20 highest tempo songs in the playlis

```{r clustering}
beast_clustering <-
  get_playlist_audio_features("Spotify", "37i9dQZF1DX76Wlfdnj7AP") %>%
  add_audio_analysis() %>%
  mutate(
    segments = map2(segments, key, compmus_c_transpose),
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "mean", norm = "manhattan"
      ),
    timbre =
      map(
        segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  ) %>%
  mutate(pitches = map(pitches, compmus_normalise, "clr")) %>%
  mutate_at(vars(pitches, timbre), map, bind_rows) %>%
  unnest(cols = c(pitches, timbre))

beast_clustering = beast_clustering[!duplicated(beast_clustering$track.name), ]
beast_clustering = beast_clustering %>%
                      arrange(desc(tempo))
beast_clustering = beast_clustering[1:20, ]

beast_juice <-
  recipe(
    track.name ~
      danceability +
      energy +
      loudness +
      speechiness +
      acousticness +
      instrumentalness +
      liveness +
      valence +
      tempo +
      duration +
      C + `C#|Db` + D + `D#|Eb` +
      E + `F` + `F#|Gb` + G +
      `G#|Ab` + A + `A#|Bb` + B +
      c01 + c02 + c03 + c04 + c05 + c06 +
      c07 + c08 + c09 + c10 + c11 + c12,
    data = beast_clustering
  ) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>% 
  # step_range(all_predictors()) %>% 
  prep(beast_clustering %>% mutate(track.name = str_trunc(track.name, 20))) %>%
  juice() %>%
  column_to_rownames("track.name")

heatmaply(
  beast_juice,
  hclustfun = hclust,
  hclust_method = "single",  # Change for single, average, or complete linkage.
  dist_method = "euclidean"
)
```

***

The dendrogram and the heatmap on the left visualize the hierarchical clusterings and feature values of the twenty highest tempo songs in the Beast Mode playlist. This dendrogram uses Euclidean distance and single linkage.\
One fascinating insight is that the tempo feature is not (almost) the same colour in every song. In contrast, these songs are the 20 highest tempo songs in the Beast Mode playlist, which has 200 songs.

There is also one clear outlier; this is Hypnocurrency. This song has a higher duration than most other songs. And has the highest tempo and instrumentalness. But because it is longer and the build-up is longer, it actually feels relatively slow. \
Certain features are too general to conclude, but we can see some similarities between songs. For example, MONTERO and GUCCI FENDI are high in acousticness and valence. And Up is high in danceability and lower in energy. At the same time, Trust nobody is lower in danceability and higher in energy.





# Conclusion {data-icon="fa-info"}

## Column 1 {data-width="500"}

### Findings and contribution

While making this portfolio, I realised there is more behind songs than I ever realised. Many features make a song in a specific genre or class, but they do not say everything. 

The expected features for a workout playlist were in most of the songs in the Beast Mode playlist. You want to have songs with high energy because these songs can change your response to fatigue and even boost your endurance. You want tracks with a high tempo that bring your workouts to the next level. Your heart rate adjusts to the beat of the music. So take this into account when you are starting your session. Are you going for a run, or are you doing strength work? Another essential feature is loudness. Loudness can help you forget the pain and let you focus on the exercises. It also helps you forget other things like emotional issues, the daily grind and other distractions.

Motivational lyrics can help you in your workout. Doing your workout and listening to music is a solo activity, so having an artist sing to you along the way can be reassuring. It can help you focus and your movements. But lyric was something that was missing in the Beast Mode playlist.

So if your a making your own workout playlist, hold into account what kind of workout you are going to do. Try to use tracks with a tempo that matches the correct heart rate for your exercises. Gyms, for example, should have tracks with a tempo between 120 and 140 BPM for strength works.\
Another thing you should consider is the loudness; you want to focus on your exercise. Forget that you have a rough day. This really can elevate your mod and spike your performance. \
Also, choose songs that you like. This sounds like an obvious choice, but this can really help you in your workout. It also helps make the workout less monotonous. 

And finally, keep some other workout playlists as a backup. It’s normal to dread a workout. Instead of forcing yourself through one playlist, build reserve playlists to match a lower energy level so you don’t feel struggling to keep up.

## Column 2 {data-width="500"}

### Reference list

Blackburn, K. B. (2019, May 8). How to get your heart rate up. MD Anderson Cancer Center. Retrieved April 3, 2022, from https://www.mdanderson.org/publications/focused-on-health/how-to-get-your-hear-rate-up.h23Z1592202.html

Buentello, S. (2021, December 28). Explore your activity on Spotify with R and "spotifyr": How to analyze and visualize your streaming history and music tastes. Medium. Retrieved April 3, 2022, from [[https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526\\\\](https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526){.uri}](%5Bhttps://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526%5D(https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526)%7B.uri%7D){.uri}

Gomez, P., & Danuser, B. (2007). Relationships between musical structure and psychophysiological measures of emotion. Emotion, 7(2), 377–387. https://doi.org/10.1037/1528-3542.7.2.377

Jones, A., & Jones, A. (2021, November 2). Umm, This May Be The Greatest Workout Playlist Of All Time. Women's Health. Retrieved April 3, 2022, from <https://www.womenshealthmag.com/fitness/a27511573/best-workout-songs-playlist/>

Pavlik, K. (2020, January 3). Classifying genres in R using Spotify data. Kaylin Pavlik. Retrieved April 3, 2022, from <https://www.kaylinpavlik.com/classifying-songs-genres/>

S., & K. (2019, November 17). How to Combine Multiple GGPlots into a Figure. Datanovia. Retrieved April 3, 2022, from https://www.datanovia.com/en/lessons/combine-multiple-ggplots-into-a-figure/

Web API Reference \| Spotify for Developers. (n.d.). Spotify. Retrieved April 3, 2022, from <https://developer.spotify.com/documentation/web-api/reference/#/operations/get-audio-features>
