---
  title: "Computational Musicology"
  author: "Melle van der Brugge"
  output:
    flexdashboard::flex_dashboard:
      orientation: columns
      vertical_layout: fill
      storyboard: true
      theme: 
        bg: "#f4b41a"
        fg: "#143d59" 
      social: menu
      source_code: embed
      navbar: 
      - { title: "Link to playlist", href: "https://open.spotify.com/playlist/37i9dQZF1DX76Wlfdnj7AP", align: right, icon: glyphicon}
      self_contained: false
---

```{r setup}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(spotifyr)
library(ggplot2)
library(shiny)
library(compmus)
library(knitr)
library(tidymodels)
library(heatmaply)
library(ggdendro)
library(ggpubr)
```

# The Playlist {data-icon="fa-music"}

**Get your Beast Mode on**!\
"*Beast mode refers to the aggressive, animalistic persona that one might assume when in competition or combat in order to overpower an opponent*."

The corpus I chose for Computational Musicology is playlists with the label "Workout". During covid, I often went to the gym to get out of the house. And if I went to the gym, I always took my phone and Air Pods with me. Music was an essential part of my gym workouts. And I always listen to the label "Workout" on Spotify during my sessions. So, what does it mean to be a "workout" playlist? One thing that interests me is that it needs to give you energy because you are exercising, and you do not want to stop. It needs to help you focus on your movements. And you don't want to get bored or feel the pain from the exercise. Another exciting thing for me is the question, "how does a "workout" playlist distinguish itself from another playlist?". Workout music often has a bumping beat, but party music can also have this, so what is the difference between party music and workout music. And does all the workout music have lyrics to keep yourself motivated, or is this not necessary. One should consider the emotions, memories, and associations that different songs evoke. For some people, the extent to which they identify with the singer's emotional state and viewpoint determines how motivated they feel

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/playlist/37i9dQZF1DX76Wlfdnj7AP?utm_source=generator" width="100%" height="380" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture">

</iframe>

The beast mode playlist is a playlist with 200 songs and has around 8.8 milion likes! The Beast mode playlist, made by Spotify, is therefore the most popular workout playlist on Spotify\
But how does a workout playlist distinguish itself from other playlists? And what are the characteristics of a workout playlist? Check out the storyboard and find out!

# Storyboard {.storyboard data-icon="fa-tachometer"}

### **Introduction**

It can be hard to stay motivated during a workout, especially when you have sooo many other things on your mind. But exercise can help distract you from your to-do lists and every day anxieties, something we all could really use during the age of the novel coronavirus.

"We know about the mental and physical benefits of exercise for the brain, and that it aids with sleep," says Jordan Metzl, MD, a sports medicine physician at Hospital for Special Surgery in New York City previously told Women's Health. "When people are ruminating about something, exercise helps."

If you're struggling to stay in the zone, there are a couple of things you can do to really amplify your sweat sesh. Lacing up a fresh pair of sneakers, for starters. Or a crowd favorite: cranking the beats.

The right workout playlist can make a huge difference in how you feel and perform during your workouts. Listening to high-tempo music (the kind that's around 170 heartbeats per minute), made people perceive their workouts as being easier, even when they were challenging, and t boosted cardiovascular benefits more than lower tempo tunes, according to a recent study published in Frontiers in Psychology.

What's more: Music may change the way that you respond to fatigue---and even boost your endurance, per research published in the Journal of Sports Exercise Psychology. Plus, other research from Sage Journal has shown that tunes can make you feel more powerful and that people who listen to music they deemed "pleasing" had higher levels of serotonin, also known as the "feel-good" hormone, according to findings published in the book Progress in Brain Research. Combine that with a workout, which already amps up your endorphins, and you'll really be treating your body right (Jones & Jones, 2021).

### **Exploring the playlist:** Understanding the Beast Mode playlist using Spotify audio features

```{r exploring}
beast_mode <-
  get_playlist_audio_features("", "37i9dQZF1DX76Wlfdnj7AP")

feature_names <- names(beast_mode)[6:16]

beast_mode %>%
  select(c('playlist_name', feature_names)) %>%
  pivot_longer(cols = feature_names) %>%
  ggplot(aes(x = value)) +
  geom_density(aes(color = playlist_name), alpha = 0.5) +
  facet_wrap(~name, ncol = 3, scales = 'free') +
  labs(title = 'Spotify Audio Feature Density',
       x = '', y = 'density') +
  theme(axis.text.y = element_blank()) 


```

------------------------------------------------------------------------

When your mission is to create the perfect workout playlist, the first thing you want is to have songs you enjoy off course. But there are some other features you want your songs to have in you workout playlist. You want ubeat and energising tracks that will keep you moving. You maybe want songs that have inspiring and motivational lyrics. And you should consider the beat; working out in time to the beat can help you get in you "zone" so you can make most of you workout.

A few features jump out in the figure on the left.\
The danceabillity of the playlist is around 0.74. A value of 0.0 is least danceable and 1.0 is most danceable. Danceabillity describes how suitable a track is for dancing based on a combination of musical elements including tempo, rythm stabillity, beat strength, and overall regularity. You also want this in your workout playlist to keep the tempo, energy in your workout

Energy is also high in the Beast mode playlist. Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. When doing you workout, you want to have the energy the finish all you exercises. Songs having a high energy will help you achieve this.

The instrumentalness is very low in the Beast mode playlist. So the predictions of Spotify are that the tracks do contain vocal content. And this is something you want in your playlist, lyrics voor inspiring and moticvation, but also nice beats to keep yopur energy

The speechiness is mostly onder 0.33, representing music and other non-speech-like tracks. This is an interesting insight in the corpus, because inspiring and motivational lyrics could help you in your workout. Apparently, people who listen to the Beast mode playlist have a preference in danceability and energy over inspiring and motivational lyrics.

The loudness of the playlist has an average at around -5. The overall loudness of a track is in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typically range between -60 and 0 db.\
Because the range is between -60 and 0,  -5 is actually quite high. But why does loudness help in your workout? Volume itself is not necessarily a contributing factor towards the level of enjoyment. It is more the fact that the louder that music is, the quieter everything else seems in comparison. One can literally drown out daily life, the daily grind, personal emotional issues etc all seem to periodically 'go away' whilst listening to loud music. It becomes a focus point, you can fully focus on your workout. There are no background noises that interrupts the flow of your gym session.

If you want to know the meaning of all Spotify audio features, check out the track's audio features: <https://developer.spotify.com/documentation/web-api/reference/#/operations/get-audio-features>

### **The Workout Feeling:** Which emotions are conveyed by the songs in Beast Mode

```{r feeling}
beast_mode <-
  get_playlist_audio_features("", "37i9dQZF1DX76Wlfdnj7AP")

emotionalQuadrant <- ggplot(data = beast_mode, aes(x = valence, y = energy, color = "red")) +
  geom_jitter() +
  geom_vline(xintercept = 0.5) +
  geom_hline(yintercept = 0.5) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 1)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  annotate('text', 0.25 / 2, 0.95, label = "Angry / Turbulent") +
  annotate('text', 1.75 / 2, 0.95, label = "Joyful / Happy") +
  annotate('text', 1.75 / 2, 0.05, label = "Peace / Chill") +
  annotate('text', 0.25 / 2, 0.05, label = "Depressing / Sad") +
  labs(x= "Valence", y= "Energy") +
  ggtitle("Emotional quadrant Beast mode", "Based on energy y valence")  
emotionalQuadrant
```

------------------------------------------------------------------------

In the figure "Emotional quadrant Beast mode" you can see what feeling the Beast mode playlist gives. Do songs convey peace, joy, anger, sadness? We can answer this question by looking at two measures Spotify has established: Valence and Energy.

Defined by Spotify, valence defines it as follows: A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry). On the other hand, energy defines it as follows: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.

With these two measurements, you can create a quadrant with a scatter plot that reveals this information. As you can see, the Beast mode playlist is high in Energy and Valence has a nice distribution.

### **The Remix Effect:** What is the difference in tempo between a "normal" version and and remixed version

```{r trust nobody}
trust_nobody_normal <- get_tidy_audio_analysis("08bDGEzsYd9ebCc7p23X4j")
      
normal <- trust_nobody_normal %>%
  tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) %>%
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

trust_nobody_remix_valentino_khan <- get_tidy_audio_analysis("52wqMbmCqeTw37jd4zFAWr")

remix <- trust_nobody_remix_valentino_khan %>%
  tempogram(window_size = 8, hop_size = 1, cyclic = TRUE) %>%
  ggplot(aes(x = time, y = bpm, fill = power)) +
  geom_raster() +
  scale_fill_viridis_c(guide = "none") +
  labs(x = "Time (s)", y = "Tempo (BPM)") +
  theme_classic()

ggarrange(normal, remix, 
          labels = c("Normal Version", "Remix"),
          ncol = 1, nrow = 2)

```

------------------------------------------------------------------------

Here you can see two tempograms, both of the song "Trust Nobody". Except, one is the "normal" version of the track and the other is the remixed version of the track. The remix version is in the Beast mode playlist, the normal version is not.

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/album/3wRs5NGdWArDIHoUw9UNMJ?utm_source=generator" width="100%" height="80" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture">

</iframe>

You can see that the beat of the song is inconsistent until around 30 second in the song. After 30 second, a line is recognizable at around 150 BPM. The song starts slower, building up the the beat drop. This repeats after around 105 seconds. So it is clearly visible that the song has two slower parts, both around 30 seconds. This can be the reason why the remix versiojn is in the Beasrt mode playlist, and not the normal version.

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/album/6r58Yml4mD6I4MXwN4UNkx?utm_source=generator" width="100%" height="80" frameBorder="0" allowfullscreen allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture">

</iframe>

In contrast to the normal version,the Valentino Khan version of Trust Nobody has a clear line almost the whole song at around 157 BPM. The BPM is even higher then the normal version, but the biggest difference is the missing of the two slower parts in the remix version. The beat is the whole song very high. And this can be the reason why this song is in the Beast mode playlist. Earlier studies reported that listening to music with a fast tempo caused an increase in the heart rate (HR) and the respiratory rate (Gomez & Danuser, 2007). And getting you heart to beat faster helps your body to move oxygen and blood to your muscles (Blackburn, 2019), helping you in your workout!



### **Structure:** Why is 'Do It To It' the most popular song

```{r self similarity features}
	
do_it_to_it <-
  get_tidy_audio_analysis("20on25jryn53hWghthWWW3") %>%
  compmus_align(bars, segments) %>%
  select(bars) %>%
  unnest(bars) %>%
  mutate(
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "acentre", norm = "manhattan"
      )
  ) %>%
  mutate(
    timbre =
      map(segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  )
bind_rows(
  do_it_to_it %>% 
    compmus_self_similarity(pitches, "aitchison") %>% 
    mutate(d = d / max(d), type = "Chroma"),
  do_it_to_it %>% 
    compmus_self_similarity(timbre, "euclidean") %>% 
    mutate(d = d / max(d), type = "Timbre")
) %>%
  mutate() %>% 
  ggplot(
    aes(
      x = xstart + xduration / 2,
      width = xduration,
      y = ystart + yduration / 2,
      height = yduration,
      fill = d
    )
  ) +
  geom_tile() +
  coord_fixed() +
  facet_wrap(~type) +
  scale_fill_viridis_c(option = "E", guide = "none") +
  theme_classic() + 
  labs(x = "", y = "Do It To It")
```

***

The two self-similarity matrices on the left, each show  pitch- and timbre-based self-similarity within the song "Do it to it" by ACRAZE and Cherish. Do It To It is the most popular track in the Beast Mode playlist. But can we understand the structure of the song and conclude why this is the most popular song, and why it is a good workout song? 

<iframe style="border-radius:12px" src="https://open.spotify.com/embed/track/20on25jryn53hWghthWWW3?utm_source=generator" width="100%" height="80" frameBorder="0" allowfullscreen="" allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture"></iframe>




### **Clustering**

```{r clustering}
beast_clustering <-
  get_playlist_audio_features("Spotify", "37i9dQZF1DX76Wlfdnj7AP") %>%
  add_audio_analysis() %>%
  mutate(
    segments = map2(segments, key, compmus_c_transpose),
    pitches =
      map(segments,
        compmus_summarise, pitches,
        method = "mean", norm = "manhattan"
      ),
    timbre =
      map(
        segments,
        compmus_summarise, timbre,
        method = "mean"
      )
  ) %>%
  mutate(pitches = map(pitches, compmus_normalise, "clr")) %>%
  mutate_at(vars(pitches, timbre), map, bind_rows) %>%
  unnest(cols = c(pitches, timbre))

beast_clustering = beast_clustering[!duplicated(beast_clustering$track.name), ]
beast_clustering = beast_clustering %>%
                      arrange(desc(tempo))
beast_clustering = beast_clustering[1:20, ]

beast_juice <-
  recipe(
    track.name ~
      danceability +
      energy +
      loudness +
      speechiness +
      acousticness +
      instrumentalness +
      liveness +
      valence +
      tempo +
      duration +
      C + `C#|Db` + D + `D#|Eb` +
      E + `F` + `F#|Gb` + G +
      `G#|Ab` + A + `A#|Bb` + B +
      c01 + c02 + c03 + c04 + c05 + c06 +
      c07 + c08 + c09 + c10 + c11 + c12,
    data = beast_clustering
  ) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>% 
  # step_range(all_predictors()) %>% 
  prep(beast_clustering %>% mutate(track.name = str_trunc(track.name, 20))) %>%
  juice() %>%
  column_to_rownames("track.name")

heatmaply(
  beast_juice,
  hclustfun = hclust,
  hclust_method = "average",  # Change for single, average, or complete linkage.
  dist_method = "euclidean"
)
```

***

info about plot

### **Conclusion**

jvhjhv

# Credits {data-icon="fa-info"}

## Column 1 {data-width="500"}

### UvA

## Column 2 {data-width="500"}

### Reference list

Blackburn, K. B. (2019, May 8). How to get your heart rate up. MD Anderson Cancer Center. Retrieved April 3, 2022, from https://www.mdanderson.org/publications/focused-on-health/how-to-get-your-hear-rate-up.h23Z1592202.html

Buentello, S. (2021, December 28). Explore your activity on Spotify with R and "spotifyr": How to analyze and visualize your streaming history and music tastes. Medium. Retrieved April 3, 2022, from [[https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526\\\\](https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526){.uri}](%5Bhttps://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526%5D(https://towardsdatascience.com/explore-your-activity-on-spotify-with-r-and-spotifyr-how-to-analyze-and-visualize-your-stream-dee41cb63526)%7B.uri%7D){.uri}

Gomez, P., & Danuser, B. (2007). Relationships between musical structure and psychophysiological measures of emotion. Emotion, 7(2), 377–387. https://doi.org/10.1037/1528-3542.7.2.377

Jones, A., & Jones, A. (2021, November 2). Umm, This May Be The Greatest Workout Playlist Of All Time. Women's Health. Retrieved April 3, 2022, from <https://www.womenshealthmag.com/fitness/a27511573/best-workout-songs-playlist/>

Pavlik, K. (2020, January 3). Classifying genres in R using Spotify data. Kaylin Pavlik. Retrieved April 3, 2022, from <https://www.kaylinpavlik.com/classifying-songs-genres/>

S., & K. (2019, November 17). How to Combine Multiple GGPlots into a Figure. Datanovia. Retrieved April 3, 2022, from https://www.datanovia.com/en/lessons/combine-multiple-ggplots-into-a-figure/

Web API Reference \| Spotify for Developers. (n.d.). Spotify. Retrieved April 3, 2022, from <https://developer.spotify.com/documentation/web-api/reference/#/operations/get-audio-features>
